canaccessR: An open data product for analyzing transportation accessibility to employment and grocery stores in Canada's largest metropolitan areas.

### 2025-03-09 paper/Manuscript-Data-Package.Rmd
describe travel time matrices
describe travel time matrices; next step is estimate descriptive statistics.

---
title: "canaccessR: An open data product for analyzing transportation accessibility to employment and grocery stores in Canada’s largest metropolitan areas."
author:
  - name: João Pedro Figueira Amorim Parga
    email: joao.parga@mail.utoronto.ca
    affil-id: 1
    orcid: 0000-0002-4105-5927
    footnote: 1
  - name: Anastasia Soukhov
    email: soukhoa@mcmaster.ca
    affil-id: 2
    orcid: 0000-0003-4371-4831
  - name: Robert Nutifafa Arku
    email: rn.arku@mail.utoronto.ca
    affil-id: 1
    orcid: 0000-0002-2018-886X
  - name: Christopher Higgins
    email: cd.higgins@utoronto.ca
    affil-id: 3
    orcid: 0000-0002-3551-7750
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affil-id: 2
    orcid: 0000-0001-6912-9919
affiliations:
  - id: 1
    name: University of Toronto
  - id: 2
    name: McMaster University
  - id: 3
    name: University of Toronto Scarborough
# address:
#   - code: CODE
#     address: "Some school"
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
  "In this paper, we describe the {canaccesR} package, an open data product (ODPs) created using the R statistical language. {canaccess} is a data package that disseminates public transit travel time estimates (travel time matrices - TTM) to employment locations and grocery stores across the 12 largest Canadian metropolitan areas. We calculate these estimates for each Dissemination Area (DA) within these regions for the years 2019 and 2023. To do so, we use the {r5r} R package, General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data. This data package can be used by researchers, practitioners, and transit agencies to estimate accessibility to essential services across these regions. These estimates can be used to compare different regions across Canada in terms of their accessibility and to conduct within-region equity assessments regarding access to services, which can inform improvements in transportation policies related to accessibility. The package is still in its initial phase and may undergo expansions in the future by adding TTM's for other destinations (e.g., schools, healthcare facilities). Finally, as an ODPs, the {canaccess} package allows for open exploration, use, and contribution by users through its GitHub repository."

@@

The objective of this paper is to describe the {canaccesR} open data package.
Its main contents are a set of public transit travel time matrices (TTM) estimates to employment and groceries stores from the 12 largest Canadian metropolitan areas in 2019 and 2023.
These estimates were created by leveraging expertise in data science, computer programming, and transportation accessibility using the {r5r} R package [@pereiraR5rRapidRealistic2021].
We used public transit schedule, transport network, population, and business location data from different sources to estimate these TTM's datasets.

Estimating accessibility - the potential offered by the transportation system to reach destinations [@paezMeasuringAccessibilityPositive2012] - requires specialized datasets and technical expertise.
Recent efforts following an open-source and transparent philosophy have been made to disseminate useful data and information on transportation in the Canadian context [@soukhovTTS2016RDataSet2023]. 
However, despite these initiatives, pre-processed and available data on transportation accessibility is scarse.
Within this context, we expect to help filling this gap by the processing raw data into user-ready data and making them publicly available to advance knowledge on the field.

To create the package, we integrated and processed raw data from diverse sources, estimated TTM's for two destinations types (*e.g.*, jobs and groceries) across the largest cities in the country, and distributed these findings through this transparent and open source data product.
Our main contribution is to provide analysis-ready data for Canada's largest cities on the topic of transportation accessibility, thus making urban analytics in the country more accessible and contributing to future research.

@@

The locations included in the data package comprise the 12 largest (population-wise) Census metropolitan areas (CMA's) based on the 2021 Canadian Census [@governmentofcanada2021CensusPopulation2021] [^1]. 
These locations are Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo, London, and Halifax.
We used four main data sources to construct the {canaccesR} data package: General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data.

[^1]: We included Oshawa, Ontario, as part of the Greater Toronto Area (GTA) because of the former's proximity to the latter. We also included Abbotsford-Mission, British Columbia, as part of the Vancouver metropolitan area because of the former's proximity to a transit station on the region’s West Coast Express commuter rail line.

<!-- ## General Transit Feed Specification -->
We manually collected and processed the GTFS files from all transit agencies within the selected CMA's to use their information on the public transit schedule in 2019 and 2023.
<!-- ## OpenStreetMaps -->
The OpenStreetMap data for the selected areas were collected through the {osmextract} package [@gilardiOsmextractDownloadImport2025].
We used OSM data from 2019 and 2023, which provided information on the areas' transit network in two points in time.
<!-- ## Census Data -->
We collected data from the 2016 Canadian Census using the {cancensus} R package [@vonbergmannCancensusPackageAccess2022] and used its information on the spatial distribution of the population and the number of workplace locations (employment) across the CMA's [@governmentofcanada2016CensusPopulation2016].
<!-- ## DMTI’s Enhanced Points of Interest -->
Finally, we gathered and cleaned the 2023 DMTI’s Enhanced Points of Interest dataset to obtain the location of the groceries stores within every urban area selected [@dmtispatialincEnhancedPointsInterest2015]. 
We filtered the locations within the DMTI dataset using the grocery stores code from the North American Industry Classification System (NAICS) and the Standard Industrial Classification (SIC).
{**THE CODE FOR THIS ESTIMATIONS AND THE TRAVEL TIME MATRICES IS ON THE transit_death_spiral github repo. 1) SHOULD WE CITE IT? 2) IS THAT A PROBLEM?**}

@@

Using the {r5r} R package, we estimated public transit travel times for two destination types, grocery stores and jobs. 
For each amenity type, we chose a likely travel time and day of the week. 
We set a 15 minutes time window and the maximum trip duration to 120 minutes. 
For groceries stores, we set the departure date to a weekend afternoon and the departure time to between 12:00 PM to 12:15 PM on April 20, 2019 and April 22, 2023.
For employment, we ran the analysis on a typical weekday morning rush-hour commute, more specifically 8:00 to 8:15 AM departure on Tuesday, April 16, 2019 and Tuesday, April 18, 2023 [^3].
In both cases, we assumed that walking was the mode of travel from origin to transit stop and from transit stop to destination.
We aggregated all the resulting travel time matrices at the Dissemination Area (DA) level, which comprise the fundamental unit of analysis in data package.
[^3]: The one exception is Quebec City, where the routing for 2019 occurs on a Saturday and Tuesday in June (instead of April) due to the GTFS data unavailability.

@@

The main contents of the {canaccessR} package are the travel time matrices estimates for all the 12 largest Canadian cities.

@@

## Descriptive statistics

Below, we present some of the basic statistics from the travel time matrices contained in the {canaccessR} package.

@@

In this paper, we describe the {canaccesR} data package, created using the {r5r} R package and transit schedule, street network, employment, and population data.
The package's main contents refers to the ready-to-use travel time matrices for public transit to reach employment and groceries stores in Canada's 12 largest urban areas.
We expect the contents of the package to be used in transportation accessibility evaluations within and across those regions.
Moreover, these datasets can be used in further equity assessments that evaluate the distribution of accessibility across space and between social groups.
Furthermore, in the spirit of open data products [@arribas-belOpenDataProductsA2021], the package can be expanded through collaboration with other researchers by, for example, including travel time matrices to other essential destinations within the DMTI's dataset (*e.g.*, schools, healthcare, etc.).
In other words, we hope that by making these datasets publicly available, future analysis can contribute to making Canada's transportation system more just and fair, considering accessibility's as the main social good of transportation [@martensTransportJusticeDesigning2016], and the inherent connection between public transit and the "right to the city" [@cogginRightTransportMoving2015].

>>>>

---
title: "canaccessR: An open data product for analyzing transportation accessibility to employment and grocery stores in Canada’s largest metropolitan areas."
author:
  - name: João Pedro Figueira Amorim Parga
    email: joao.parga@mail.utoronto.ca
    affil-id: 1
    orcid: 0000-0002-4105-5927
    footnote: 1
  - name: Anastasia Soukhov
    email: soukhoa@mcmaster.ca
    affil-id: 2
    orcid: 0000-0003-4371-4831
  - name: Robert Nutifafa Arku
    email: rn.arku@mail.utoronto.ca
    affil-id: 1
    orcid: 0000-0002-2018-886X
  - name: Christopher Higgins
    email: cd.higgins@utoronto.ca
    affil-id: 3
    orcid: 0000-0002-3551-7750
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affil-id: 2
    orcid: 0000-0001-6912-9919
affiliations:
  - id: 1
    name: University of Toronto
  - id: 2
    name: McMaster University
  - id: 3
    name: University of Toronto Scarborough
# address:
#   - code: CODE
#     address: "Some school"
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
  "In this paper, we describe the {canaccesR} package, an open data product (ODP) created in R that contains public transit travel time estimates to employment locations and grocery stores across Canada's 12 largest metropolitan areas. We calculate travel time matrices (TTM) from and to each Dissemination Area (DA) within these regions for the years 2019 and 2023. We add value to the urban analytics community by processing and integrating raw data, and disseminating user-ready data in the domain of transportation accessibility in Canada. To do so, we use the {r5r} R package, General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data. This data package can be used by researchers, practitioners, and transit agencies to estimate accessibility levels to these two essential destinations within these urban areas. Moreover, these estimations can be used as inputs in equity and inequalities assessments, through the comparisons of within and across accessibility levels found throughout Canada's largest metropolitan areas. Consequently, we expect to contribute to informed and data-based decision making in transportation by disseminating these data. We hope that these datasets can substantiate future improvements in policy-making that may lead to greater justice in the country's urban transportation systems. The package is still in its initial phase and may undergo expansions in the future by adding TTM's for other destinations (e.g., schools, healthcare facilities). Finally, as an ODPs, the {canaccess} package allows for open exploration, use, and contribution by users through its GitHub repository."

@@

The objective of this paper is to describe the {canaccesR} open data package.
Its main contents are a set of public transit travel time matrices (TTM) estimates to employment and groceries stores from the 12 largest Canadian metropolitan areas in 2019 and 2023.
The results refer to all the Dissemination Areas (DA) (the smallest publicly available geography provided by Statistics Canada [@governmentofcanadaDictionaryCensusPopulation2021a]) within each metropolitan area, yielding origin-destination pairs containing information on travel time by public transit, population, and total employment within those spatial units.
These estimates were created by leveraging expertise in data science, computer programming, and transportation accessibility using the {r5r} package [@pereiraR5rRapidRealistic2021].
We used public transit schedule, transport network, population, and business location data from different sources to estimate these TTM's datasets.

Estimating accessibility - the potential offered by the transportation system to reach destinations [@paezMeasuringAccessibilityPositive2012] - requires specialized datasets and technical expertise.
Recent efforts following an open-source and transparent philosophy have been made to disseminate useful data and information on transportation in the Canadian context [@soukhovTTS2016RDataSet2023]. 
However, despite these initiatives, pre-processed and available data that allow for the ease estimation of accessibility indicators are still scarce.
Within this context, we expect to help filling this gap by the processing raw into user-ready data and making them publicly available to advance knowledge on the field.
Our main contribution is to provide analysis-ready data for Canada's largest cities on the topic of transportation accessibility, thus making urban analytics in the country more accessible and contributing to future research and data-based decision making.

@@

The locations included in the data package comprise the 12 largest (population-wise) Census metropolitan areas (CMA's) based on the 2021 Canadian Census [@governmentofcanada2021CensusPopulation2021] [^2]. 
These locations are Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo, London, and Halifax.
We used four main data sources to construct the {canaccesR} data package: General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data.

[^2]: We included Oshawa, Ontario, as part of the Greater Toronto Area (GTA) because of the former's proximity to the latter. We also included Abbotsford-Mission, British Columbia, as part of the Vancouver metropolitan area because of the former's proximity to a transit station on the region’s West Coast Express commuter rail line.

<!-- ## General Transit Feed Specification -->
We manually collected and processed the GTFS files from all transit agencies within the selected CMA's to use their information on the public transit schedule in 2019 and 2023.
<!-- ## OpenStreetMaps -->
The OpenStreetMap data for the selected areas were collected through the {osmextract} package [@gilardiOsmextractDownloadImport2025].
We used OSM data from 2019 and 2023, which provided information on the areas' transit network in two points in time.
<!-- ## Census Data -->
We collected data from the 2016 Canadian Census using the {cancensus} package [@vonbergmannCancensusPackageAccess2022] and used its information on the spatial distribution of the population and the number of workplace locations (employment) across the CMA's [@governmentofcanada2016CensusPopulation2016].
<!-- ## DMTI’s Enhanced Points of Interest -->
Finally, we gathered and cleaned the 2023 DMTI’s Enhanced Points of Interest dataset to obtain the location of the groceries stores within every urban area selected [@dmtispatialincEnhancedPointsInterest2015]. 
We filtered the locations within the DMTI dataset using the grocery stores code from the North American Industry Classification System (NAICS) and the Standard Industrial Classification (SIC).
{**THE CODE FOR THIS ESTIMATIONS AND THE TRAVEL TIME MATRICES IS ON THE transit_death_spiral github repo. 1) SHOULD WE CITE IT? 2) IS THAT A PROBLEM?**}

@@

Using the {r5r} package, we estimated public transit travel times for two destination types, grocery stores and jobs. 
For each amenity type, we chose a likely travel time and day of the week. 
We set a 15 minutes time window and the maximum trip duration to 120 minutes. 
The estimated times are the median of the 15 minute time window.
For groceries stores, we set the departure date to a weekend afternoon and the departure time to between 12:00 PM to 12:15 PM on April 20, 2019 and April 22, 2023.
For employment, we ran the analysis on a typical weekday morning rush-hour commute, more specifically 8:00 to 8:15 AM departure on Tuesday, April 16, 2019 and Tuesday, April 18, 2023 [^3].
In both cases, we assumed that walking was the mode of travel from origin to transit stop and from transit stop to destination.
We aggregated all the resulting travel time matrices at the Dissemination Area (DA) level, which comprise the fundamental unit of analysis in data package.

[^3]: The one exception is Quebec City, where the routing for 2019 occurs on a Saturday and Tuesday in June (instead of April) due to the GTFS data unavailability.

@@

## Travel time matrices

The main contents of the {canaccessR} package are the travel time matrices.
<!-- These matrices are data frames with origin-destination pairs, travel time by public transit, and population and employment by Census Dissemination Area.  -->
These matrices (data frames) comprise the estimated travel time by transit from each origin and destination Dissemination Area pairs.
The matrices also contain identifiers, such as the unique origin and destination DA codes, region identifier (*e.g.*, "cal" for the region Calgary) and name, the year in which the travel times refers to, and the date and time of departure for a trip that originates at the Dissemination Area of origin.
Besides these information, the matrices feature the total population in the origin DA and the total employment in the destination DA (both according to the 2016 Census).
There are two matrices for each metropolitan area, one containing the travel times to jobs and the other to groceries stores, making it 20 datasets (the information from Toronto, Hamilton, and Waterloo are aggregated at the Greater Golden Horseshoe travel time matrices).

**HOW TO MAKE THE INFORMATION FROM THE TTM INTO SPATIAL DATA?**
The information can be turned into spatial data using the

Below, we present some descriptive statistics from the travel time matrices contained in the {canaccessR} package.

## Other contents

@@

@@

In this paper, we describe the {canaccesR} data package, created using the {r5r} package and transit schedule, street network, employment, and population data.
The package's main contents refers to the ready-to-use travel time matrices for public transit to reach employment and groceries stores in Canada's 12 largest urban areas.
We expect the contents of the package to be used in transportation accessibility evaluations within and across those regions.
Moreover, these datasets can be used in further equity assessments that evaluate the distribution of accessibility across space and between social groups.
Furthermore, in the spirit of open data products [@arribas-belOpenDataProductsA2021], the package can be expanded through collaboration with other researchers by, for example, including travel time matrices to other essential destinations within the DMTI's dataset (*e.g.*, schools, healthcare, etc.).
In other words, we hope that by making these datasets publicly available, future analysis can contribute to making Canada's transportation system more just and fair, considering accessibility's as the main social good of transportation [@martensTransportJusticeDesigning2016], and the inherent connection between public transit and the "right to the city" [@cogginRightTransportMoving2015].

### 2025-03-09 paper/Manuscript-Data-Package.tex
describe travel time matrices
describe travel time matrices; next step is estimate descriptive statistics.

\begin{abstract}
In this paper, we describe the \{canaccesR\} package, an open data
product (ODPs) created using the R statistical language. \{canaccess\}
is a data package that disseminates public transit travel time estimates
(travel time matrices - TTM) to employment locations and grocery stores
across the 12 largest Canadian metropolitan areas. We calculate these
estimates for each Dissemination Area (DA) within these regions for the
years 2019 and 2023. To do so, we use the \{r5r\} R package, General
Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI's Enhanced
Points of Interest, and Statistics Canada Census data. This data package
can be used by researchers, practitioners, and transit agencies to
estimate accessibility to essential services across these regions. These
estimates can be used to compare different regions across Canada in
terms of their accessibility and to conduct within-region equity
assessments regarding access to services, which can inform improvements
in transportation policies related to accessibility. The package is
still in its initial phase and may undergo expansions in the future by
adding TTM's for other destinations (e.g., schools, healthcare
facilities). Finally, as an ODPs, the \{canaccess\} package allows for
open exploration, use, and contribution by users through its GitHub
repository.
\end{abstract}

@@

The objective of this paper is to describe the \{canaccesR\} open data
package. Its main contents are a set of public transit travel time
matrices (TTM) estimates to employment and groceries stores from the 12
largest Canadian metropolitan areas in 2019 and 2023. These estimates
were created by leveraging expertise in data science, computer
programming, and transportation accessibility using the \{r5r\} R
package \citep{pereiraR5rRapidRealistic2021}. We used public transit
schedule, transport network, population, and business location data from
different sources to estimate these TTM's datasets.

Estimating accessibility - the potential offered by the transportation
system to reach destinations
\citep{paezMeasuringAccessibilityPositive2012} - requires specialized
datasets and technical expertise. Recent efforts following an
open-source and transparent philosophy have been made to disseminate
useful data and information on transportation in the Canadian context
\citep{soukhovTTS2016RDataSet2023}. However, despite these initiatives,
pre-processed and available data on transportation accessibility is
scarse. Within this context, we expect to help filling this gap by the
processing raw data into user-ready data and making them publicly
available to advance knowledge on the field.

To create the package, we integrated and processed raw data from diverse
sources, estimated TTM's for two destinations types (\emph{e.g.}, jobs
and groceries) across the largest cities in the country, and distributed
these findings through this transparent and open source data product.
Our main contribution is to provide analysis-ready data for Canada's
largest cities on the topic of transportation accessibility, thus making
urban analytics in the country more accessible and contributing to
future research.

@@

We manually collected and processed the GTFS files from all transit
agencies within the selected CMA's to use their information on the
public transit schedule in 2019 and 2023. The OpenStreetMap data for the
selected areas were collected through the \{osmextract\} package
\citep{gilardiOsmextractDownloadImport2025}. We used OSM data from 2019
and 2023, which provided information on the areas' transit network in
two points in time. We collected data from the 2016 Canadian Census
using the \{cancensus\} R package
\citep{vonbergmannCancensusPackageAccess2022} and used its information
on the spatial distribution of the population and the number of
workplace locations (employment) across the CMA's
\citep{governmentofcanada2016CensusPopulation2016}. Finally, we gathered
and cleaned the 2023 DMTI's Enhanced Points of Interest dataset to
obtain the location of the groceries stores within every urban area
selected \citep{dmtispatialincEnhancedPointsInterest2015}. We filtered
the locations within the DMTI dataset using the grocery stores code from
the North American Industry Classification System (NAICS) and the
Standard Industrial Classification (SIC). \{\textbf{THE CODE FOR THIS
ESTIMATIONS AND THE TRAVEL TIME MATRICES IS ON THE
transit\_death\_spiral github repo. 1) SHOULD WE CITE IT? 2) IS THAT A
PROBLEM?}\}

@@

Using the \{r5r\} R package, we estimated public transit travel times
for two destination types, grocery stores and jobs. For each amenity
type, we chose a likely travel time and day of the week. We set a 15
minutes time window and the maximum trip duration to 120 minutes. For
groceries stores, we set the departure date to a weekend afternoon and
the departure time to between 12:00 PM to 12:15 PM on April 20, 2019 and
April 22, 2023. For employment, we ran the analysis on a typical weekday
morning rush-hour commute, more specifically 8:00 to 8:15 AM departure
on Tuesday, April 16, 2019 and Tuesday, April 18, 2023 {[}\^{}3{]}. In
both cases, we assumed that walking was the mode of travel from origin
to transit stop and from transit stop to destination. We aggregated all
the resulting travel time matrices at the Dissemination Area (DA) level,
which comprise the fundamental unit of analysis in data package.
{[}\^{}3{]}: The one exception is Quebec City, where the routing for
2019 occurs on a Saturday and Tuesday in June (instead of April) due to
the GTFS data unavailability.

@@

The main contents of the \{canaccessR\} package are the travel time
matrices estimates for all the 12 largest Canadian cities.

@@

\subsection{Descriptive statistics}\label{descriptive-statistics}

Below, we present some of the basic statistics from the travel time
matrices contained in the \{canaccessR\} package.

@@

In this paper, we describe the \{canaccesR\} data package, created using
the \{r5r\} R package and transit schedule, street network, employment,
and population data. The package's main contents refers to the
ready-to-use travel time matrices for public transit to reach employment
and groceries stores in Canada's 12 largest urban areas. We expect the
contents of the package to be used in transportation accessibility
evaluations within and across those regions. Moreover, these datasets
can be used in further equity assessments that evaluate the distribution
of accessibility across space and between social groups. Furthermore, in
the spirit of open data products
\citep{arribas-belOpenDataProductsA2021}, the package can be expanded
through collaboration with other researchers by, for example, including
travel time matrices to other essential destinations within the DMTI's
dataset (\emph{e.g.}, schools, healthcare, etc.). In other words, we
hope that by making these datasets publicly available, future analysis
can contribute to making Canada's transportation system more just and
fair, considering accessibility's as the main social good of
transportation \citep{martensTransportJusticeDesigning2016}, and the
inherent connection between public transit and the ``right to the city''
\citep{cogginRightTransportMoving2015}.

>>>>

\begin{abstract}
In this paper, we describe the \{canaccesR\} package, an open data
product (ODP) created in R that contains public transit travel time
estimates to employment locations and grocery stores across Canada's 12
largest metropolitan areas. We calculate travel time matrices (TTM) from
and to each Dissemination Area (DA) within these regions for the years
2019 and 2023. We add value to the urban analytics community by
processing and integrating raw data, and disseminating user-ready data
in the domain of transportation accessibility in Canada. To do so, we
use the \{r5r\} R package, General Transit Feed Specification (GTFS),
OpenStreetMap (OSM), DMTI's Enhanced Points of Interest, and Statistics
Canada Census data. This data package can be used by researchers,
practitioners, and transit agencies to estimate accessibility levels to
these two essential destinations within these urban areas. Moreover,
these estimations can be used as inputs in equity and inequalities
assessments, through the comparisons of within and across accessibility
levels found throughout Canada's largest metropolitan areas.
Consequently, we expect to contribute to informed and data-based
decision making in transportation by disseminating these data. We hope
that these datasets can substantiate future improvements in
policy-making that may lead to greater justice in the country's urban
transportation systems. The package is still in its initial phase and
may undergo expansions in the future by adding TTM's for other
destinations (e.g., schools, healthcare facilities). Finally, as an
ODPs, the \{canaccess\} package allows for open exploration, use, and
contribution by users through its GitHub repository.
\end{abstract}

@@

The objective of this paper is to describe the \{canaccesR\} open data
package. Its main contents are a set of public transit travel time
matrices (TTM) estimates to employment and groceries stores from the 12
largest Canadian metropolitan areas in 2019 and 2023. The results refer
to all the Dissemination Areas (DA) (the smallest publicly available
geography provided by Statistics Canada
\citep{governmentofcanadaDictionaryCensusPopulation2021a}) within each
metropolitan area, yielding origin-destination pairs containing
information on travel time by public transit, population, and total
employment within those spatial units. These estimates were created by
leveraging expertise in data science, computer programming, and
transportation accessibility using the \{r5r\} package
\citep{pereiraR5rRapidRealistic2021}. We used public transit schedule,
transport network, population, and business location data from different
sources to estimate these TTM's datasets.

Estimating accessibility - the potential offered by the transportation
system to reach destinations
\citep{paezMeasuringAccessibilityPositive2012} - requires specialized
datasets and technical expertise. Recent efforts following an
open-source and transparent philosophy have been made to disseminate
useful data and information on transportation in the Canadian context
\citep{soukhovTTS2016RDataSet2023}. However, despite these initiatives,
pre-processed and available data that allow for the ease estimation of
accessibility indicators are still scarce. Within this context, we
expect to help filling this gap by the processing raw into user-ready
data and making them publicly available to advance knowledge on the
field. Our main contribution is to provide analysis-ready data for
Canada's largest cities on the topic of transportation accessibility,
thus making urban analytics in the country more accessible and
contributing to future research and data-based decision making.

@@

We manually collected and processed the GTFS files from all transit
agencies within the selected CMA's to use their information on the
public transit schedule in 2019 and 2023. The OpenStreetMap data for the
selected areas were collected through the \{osmextract\} package
\citep{gilardiOsmextractDownloadImport2025}. We used OSM data from 2019
and 2023, which provided information on the areas' transit network in
two points in time. We collected data from the 2016 Canadian Census
using the \{cancensus\} package
\citep{vonbergmannCancensusPackageAccess2022} and used its information
on the spatial distribution of the population and the number of
workplace locations (employment) across the CMA's
\citep{governmentofcanada2016CensusPopulation2016}. Finally, we gathered
and cleaned the 2023 DMTI's Enhanced Points of Interest dataset to
obtain the location of the groceries stores within every urban area
selected \citep{dmtispatialincEnhancedPointsInterest2015}. We filtered
the locations within the DMTI dataset using the grocery stores code from
the North American Industry Classification System (NAICS) and the
Standard Industrial Classification (SIC). \{\textbf{THE CODE FOR THIS
ESTIMATIONS AND THE TRAVEL TIME MATRICES IS ON THE
transit\_death\_spiral github repo. 1) SHOULD WE CITE IT? 2) IS THAT A
PROBLEM?}\}

@@

Using the \{r5r\} package, we estimated public transit travel times for
two destination types, grocery stores and jobs. For each amenity type,
we chose a likely travel time and day of the week. We set a 15 minutes
time window and the maximum trip duration to 120 minutes. The estimated
times are the median of the 15 minute time window. For groceries stores,
we set the departure date to a weekend afternoon and the departure time
to between 12:00 PM to 12:15 PM on April 20, 2019 and April 22, 2023.
For employment, we ran the analysis on a typical weekday morning
rush-hour commute, more specifically 8:00 to 8:15 AM departure on
Tuesday, April 16, 2019 and Tuesday, April 18, 2023 \footnote{The one
  exception is Quebec City, where the routing for 2019 occurs on a
  Saturday and Tuesday in June (instead of April) due to the GTFS data
  unavailability.}. In both cases, we assumed that walking was the mode
of travel from origin to transit stop and from transit stop to
destination. We aggregated all the resulting travel time matrices at the
Dissemination Area (DA) level, which comprise the fundamental unit of
analysis in data package.

@@

\subsection{Travel time matrices}\label{travel-time-matrices}

The main contents of the \{canaccessR\} package are the travel time
matrices. These matrices (data frames) comprise the estimated travel
time by transit from each origin and destination Dissemination Area
pairs. The matrices also contain identifiers, such as the unique origin
and destination DA codes, region identifier (\emph{e.g.}, ``cal'' for
the region Calgary) and name, the year in which the travel times refers
to, and the date and time of departure for a trip that originates at the
Dissemination Area of origin. Besides these information, the matrices
feature the total population in the origin DA and the total employment
in the destination DA (both according to the 2016 Census). There are two
matrices for each metropolitan area, one containing the travel times to
jobs and the other to groceries stores, making it 20 datasets (the
information from Toronto, Hamilton, and Waterloo are aggregated at the
Greater Golden Horseshoe travel time matrices).

\textbf{HOW TO MAKE THE INFORMATION FROM THE TTM INTO SPATIAL DATA?} The
information can be turned into spatial data using the

Below, we present some descriptive statistics from the travel time
matrices contained in the \{canaccessR\} package.

\subsection{Other contents}\label{other-contents}

@@

@@

In this paper, we describe the \{canaccesR\} data package, created using
the \{r5r\} package and transit schedule, street network, employment,
and population data. The package's main contents refers to the
ready-to-use travel time matrices for public transit to reach employment
and groceries stores in Canada's 12 largest urban areas. We expect the
contents of the package to be used in transportation accessibility
evaluations within and across those regions. Moreover, these datasets
can be used in further equity assessments that evaluate the distribution
of accessibility across space and between social groups. Furthermore, in
the spirit of open data products
\citep{arribas-belOpenDataProductsA2021}, the package can be expanded
through collaboration with other researchers by, for example, including
travel time matrices to other essential destinations within the DMTI's
dataset (\emph{e.g.}, schools, healthcare, etc.). In other words, we
hope that by making these datasets publicly available, future analysis
can contribute to making Canada's transportation system more just and
fair, considering accessibility's as the main social good of
transportation \citep{martensTransportJusticeDesigning2016}, and the
inherent connection between public transit and the ``right to the city''
\citep{cogginRightTransportMoving2015}.
